A Fast and Tight Heuristic for A* in Road Networks

### 2020-08-22 paper/ch_potentials.tex
foo

\usepackage{booktabs}
\usepackage[algo2e,vlined]{algorithm2e}

@@

\subsection{Oracle-A* \& Comparison with Related Work}

@@

In the preprocessing phase, we compute the subgraph $G_C$, called \emph{core}, induced by the largest biconnected component of the undirected graph underlying $G$.
We do this using Tarjan's algorithm \cite{t-dfslg2-72}.
For every node $v$ in the input graph $G$, we store the attachment node $a_v$ to the core.
For nodes in the core, $a_v=v$.
We exploit that all attachment nodes are single node separators and the problem can be decomposed along them.

@@

\subparagraph{Inputs and Methodology}
Our main benchmark instance is a graph of the road network of Germany obtained from Open Street Map\footnote{\url{https://download.geofabrik.de/europe/germany-200101.osm.pbf}}.
To obtain the routing graph, we use the import from RoutingKit\footnote{\url{https://github.com/RoutingKit/RoutingKit}}.
The graph has 16M nodes and 35M edges.
For this instance, we have proprietary traffic data provided by Mapbox\footnote{\url{https://www.mapbox.com}}.
The  data includes a live traffic snapshot from Friday 2019/08/02 afternoon and comes in the form of 320K OSM node pairs and live speeds for the edge between the nodes.
It also includes traffic predictions for 38\% of the edges as predicted speeds for all five minute periods over the course of a week.
We exclude speed values which are faster than the freeflow speed computed by RoutingKit.
Additionally, have two graphs with traffic predictions provided by PTV\footnote{\url{https://ptvgroup.com}}.
One is an old instance of Germany with traffic predictions from 2006 for 7\% of the edges and the other one a newer instance of Europe with predictions for 27\% of the edges.
Table~\ref{tab:graphs} contains an overview over our instances.
We report preprocessing running times as averages over 10 runs.
For queries, we perform 10\,000 point-to-point queries where both source and target are nodes drawn uniformly at random and report average results.

\subparagraph{Experiments}
We first investigate the effectiveness of the heuristic depending on how tight the lower bound is.
For that, we take our main graph, use the unmodified travel time as the lower bound $w_\ell$, and perform queries on $w_q = \alpha \cdot w_\ell$ where $\alpha$ is a factor $\geq 1$.
When $\alpha = 1$, the heuristic is as tight as possible.
With growing $\alpha$ the heuristic degrades.
Figure~\ref{fig:scaled_weights} depicts the results.
Clearly, $\alpha$ has significant influence on the running time.
Average running times range from below a millisecond to a few hundred milliseconds depending on $\alpha$.
Up to around $\alpha = 1.1$ the running time growth is roughly quadratic before leveling off to something more resembling linear growth.
We observe a significant amount of heavy outliers -- sometimes more than an order of magnitude slower than the median.
Generally, running times vary quite strongly even though most queries are long range queries as we draw source and target node uniformly at random.
The reason is that different queries (even with similar shortest path lengths) are affected differently by $\alpha$ depending on the geographic regions the shortest path traverses.
As $\alpha$ grows, more nodes in the area around the shortest paths are explored.
So when a shortest path traverses an area where many nodes are reachable in short distance, the search space grows stronger, than when it passes sparse areas.
% Conclusions?

In Table~\ref{tab:building_blocks} we investigate the impact of our optimizations.
We run queries with different heuristic functions: the \emph{Zero} heuristic which turns A* back into the classical algorithm of Dijkstra, our new CH-Potentials heuristic, and an \emph{Oracle} heuristic where all distances to $t$ with regards to $w_\ell$ can be obtained immediately.
We first observe that each one of our optimizations gains useful speedups regardless of the heuristic used.
Our optimizations do not only reduce heuristic evaluations but also queue operations which always accelerates the algorithm.
The number of queue pushs roughly corresponds to the running time for all heuristics.
When comparing CH-Potentials to the Oracle heuristic we note that CH-Potentials are only a factor of 1.6 slower.
This clearly shows that CH-Potentials have very little overhead.
Assuming that $w_\ell$ is the tightest possible lower bound that can be known in advance, no other heuristic can be more than a factor of 1.6 faster than CH-Potentials without sacrificing exactness.

Table~\ref{tab:applications} depicts results when applying our algorithm to different route planning problems.
The greatest speedup is, of course, achieved when applying CH-Potentials to an unmodified graph with $w_q = w_\ell$.
In that case, the performance is within an order of magnitude to what other state-of-the-art techniques achieve.
When considering extended problems, the performance of CH-Potentials strongly depends on the tightness of $w_\ell$.
On one hand, forbidding highways leads to over 40\% longer paths and the smallest speedup (only 5.3) among all settings.
On the other hand, incorporating turn restrictions increases path lengths only marginally and CH-Potentials yield a speedup of around 1500 over Dijkstra on the turn expanded graph without any additional implementation complexity.
Live traffic increases shortest paths lengths by around 15\% longer which yields running times of 127\,ms on average.
The Mapbox traffic predictions increase shortest path lengths by about 18\%.
Combined with the additional work from function evaluations, this leads to running times around 200\,ms.
The PTV traffic predictions yield much smaller length increases (less than 4\%) which leads to better running times and speedups.
% Assessing which of these predictions is more realistic is not possible within the scope of this paper.
Combining Mapbox live traffic and traffic predictions yields running times only slightly slower than traffic predictions alone.
% To the best of our knowledge, no other exact technique so far could combine these two scenarios.
Additionally adding turn restrictions slows running times down to around 510\,ms despite the length increase only growing by 1.3 percentage points.
The reason for this is that the turn expanded graph is three times bigger and that the BCC optimization is not applicable for turn-expanded graphs.
This is still a speedup of 12.2 over the baseline.
% To the best of our knowledge no other approach so far did even attempt handling this combination of scenarios.

@@

In this paper, we introduced CH-Potentials, a fast and exact new A* heuristic for finding shortest paths in road networks.
The approach is flexible and can handle a multitude of routing scenarios with very little implementation complexity.
The performance of CH-Potentials depends on the tightness of the lower bound, so whether CH-Potentials are fit for a certain application depends on whether a tight lower bound can be derived.
Our experiments show that CH-Potentials are efficient in extracting distance estimates from a lower bound graph and only around 1.6 times slower than immediately obtaining these estimates from an hypothetical oracle.

>>>>

\usepackage{booktabs}
\usepackage{multirow}
\usepackage[algo2e,vlined]{algorithm2e}

@@

\subsection{Oracle-A* \& Comparison with Related Work}
\label{sec:oracle}

@@

In the preprocessing phase, we compute the subgraph $G_C$, called \emph{core}.
$G_C$ is induced by the largest biconnected component of the undirected graph underlying $G$.
We do this using Tarjan's algorithm \cite{t-dfslg2-72}.
For every node $v$ in the input graph $G$, we store the attachment node $a_v$ to the core.
For nodes in the core, $a_v=v$.
We exploit that all attachment nodes are single node separators and the problem can be decomposed along them.

@@

\subparagraph{Inputs and Methodology}
Our main benchmark instance is a graph of the road network of Germany obtained from Open Street Map\footnote{\url{download.geofabrik.de/europe/germany-200101.osm.pbf}}.
To obtain the routing graph, we use the import from RoutingKit\footnote{\url{github.com/RoutingKit/RoutingKit}}.
The graph has 16M nodes and 35M edges.
For this instance, we have proprietary traffic data provided by Mapbox\footnote{\url{www.mapbox.com}}.
The  data includes a live traffic snapshot from Friday 2019/08/02 afternoon and comes in the form of 320K OSM node pairs and live speeds for the edge between the nodes.
It also includes traffic predictions for 38\% of the edges as predicted speeds for all five minute periods over the course of a week.
We exclude speed values which are faster than the freeflow speed computed by RoutingKit.
Additionally, we have two graphs with proprietary traffic predictions provided by PTV\footnote{\url{ptvgroup.com}}.
The PTV instances are not OSM-based.
One is an old instance of Germany with traffic predictions from 2006 for 7\% of the edges and the other one a newer instance of Europe with predictions for 27\% of the edges.
Table~\ref{tab:graphs} contains an overview over our instances.
In this table, we further include the sequential running time necessary to construct the CH.
We report preprocessing running times as averages over 10 runs.
For queries, we perform 10\,000 point-to-point queries where both source and target are nodes drawn uniformly at random and report average results.

\subparagraph{Experiments}
The first experiment varies the tightness of the heuristic.
The lower bound $w_\ell$ is set to the freeflow travel time.
The query weights $w_q$ are set to $\alpha \cdot w_\ell$, where $\alpha\ge 1$.
When $\alpha = 1$, the heuristic is tight.
Increasing $\alpha$ degrades the heuristic's quality.
Figure~\ref{fig:scaled_weights} depicts the results.
Clearly, $\alpha$ has significant influence on the running time.
Average running times range from below a millisecond to a few hundred milliseconds depending on $\alpha$.
Up to around $\alpha = 1.1$ the running time grows quickly.
After $\alpha = 1.1$ the growth slows down.
We observe that the running times for a fixed $\alpha$ vary strongly.
This is an interesting observation, as with uniform source and target sampling, nearly all queries are long-distance.
The query distance is thus not the reason.
After some investigation, we concluded that this is due to non-uniform road graph density.
Some regions have more roads per area than others.
The search space of A* is an ellipsoid.
The number explored A* nodes depends on the density of the area covered by this shape.
As the density varies, the running times vary.

Table~\ref{tab:building_blocks} shows the impact of our optimizations.
We run queries with three heuristic.
The Zero heuristic is simply $h(x)=0$ for all $x$.
This corresponds to using Dijkstra's algorithm.
Further, we run queries with our novel CH-Potential heuristic.
Finally, we use the unattainable lower bound Oracle-A* as described in Section~\ref{sec:oracle}.
Oracle-A* is unbeatable theoretic lower bound.
The reported running time of Oracle-A* do \emph{not} account for the heuristic evaluation.
We include Oracle-A* as it is a lower bound not only for our algorithm but every related work.

We observe that the number of queue pushes roughly correlates with running time.
Further, each of our optimizations reduces the number of queue pushes.
Very interesting is the comparison with Oracle-A*.
Our algorithm is only 1.6 times slower than an unachievable theoretic lower bound.
This means that no competitor algorithm such as ALT or CPD-Heuristics can be significantly faster than CH-Potentials.

Table~\ref{tab:applications} depicts the running times of CH-Potentials in the various applications, such as those described in Section~\ref{sec:extensions}.
Speedups compared to extensions of Dijkstra's algorithm for the respective application are depicted.
We start with the inflexible routing problem where $w_q = w_\ell$.
This is the only problem variant solved by the basic CH algorithm.
In this setting, there is a huge speedup of 3243.
Such large speedups are typical for CH.
This shows that CH-Potentials gracefully converges to a CH in the $w_q = w_\ell$ special case.

In the other scenarios, the performance of CH-Potentials strongly depends on the quality of the tight heuristic.
We measure this quality using the length increase of $w_q$ compared to $w_\ell$.
Forbidding highways results in the largest length increase and in the smallest speedup.
The other extreme are turn restrictions.
They have only a small impact on the length increase.
The achieved speedups are therefore comparable to CH speedups.
%
Mapbox live traffic has a length increase of around 15\%, which yields running times of 127\,ms.
The length increase of Mapbox traffic predictions are about 18\%, and results in a running time of 200\,ms.
The speedup in the predicted scenario is larger than in the live setting, as the travel time function evaluations slow down Dijkstra's algorithm.
Combining predicted and live traffic results in a running time only slightly higher than for the predicted scenario.
Further adding turn restrictions, increases the running times.
This increase is mostly due to the BCC optimization becoming ineffective when considering turns.
It is not due to the length increase of using turns.
With everything activated, our algorithm still has a speedup of 12.2 over the baseline.
Interestingly, the PTV traffic predictions have a much smaller length increase that the Mapbox predictions.
This results in smaller running times of our algorithm.

@@

In this paper, we introduced CH-Potentials, a fast, exact, and flexible two-phase algorithm based on A* and CH for finding shortest paths in road networks.
The approach can handle a multitude of routing scenarios with very little implementation complexity.
The performance of CH-Potentials crucially depends on the availability of good weight lower bounds in preprocessing phase.
Our experiments show, that this availability is highly application dependent.
We experimentally show that the running time of our algorithm is within a factor 1.6 of a hypothetical A* variant that can instantly evaluate the heuristic.
We conclude from this, that to achieve smaller running times with A* based algorithms variations of the problem setting are necessary.
This leads to multiple avenues for future research.

Dropping the provable exactness requirement using a setup similar to anytime A* \cite{DBLP:conf/aaai/ZhouH02,DBLP:conf/nips/LikhachevGT03} would be interesting.
Another promising research avenue would be to investigate other graphs than road networks.
A lot of research into grid maps exists including a series of competitions called GPPC \cite{DBLP:conf/socs/SturtevantTTUKS15}.
Hierarchical techniques have been shown to work on these graphs \cite{DBLP:conf/aaai/UrasK14}.
